---
title: "Lab_3_OLEA"
author: "Julian Olea"
format: 
  html:
    embed-resources: true
fig-width: 8
fig-height: 6
warning: FALSE
---

```{r}

setwd("~/Masters/PM_566/labs/Lab_03")

```


```{r Packages Used}

# library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(rstatix)
library(ggpubr)
library(leaflet)
library(lubridate)

```

#1: Read the data

```{r}

# met <- read_csv("met_all.gz")

met <- read.table("met_all.gz", header = TRUE, sep = ",")

```

#2: Check the dimensions, headers, footers
```{r}

dim(met)
head(met, 3)
tail(met, 3)

```
There are 2377343 and 30 columns

#4:Take a closer look at the key variables.
```{r}

table(met$year)
table(met$day)
table(met$hour)
summary(met$temp)
summary(met$elev)
summary(met$wind.sp)

```

```{r}

# met[met$elev == 9999.0 , ] <- NA

met %>% filter(!elev == 9999.0) -> met_filtered

summary(met_filtered$elev)

```
After subsetting the data to exclude the sites at the 9999 meter elevation, the new max site elevation is 4113 meters.

```{r}

met_filtered %>% filter(temp > -40) %>% select(temp) %>% summary()

```
The minimum temperature reported is -17.20 celsius, we will check if this makes sense.

#5: Check the data against an external data source.

```{r}

met_filtered %>% arrange(temp) %>% select(temp, lat, lon, elev) %>% distinct() %>% head(3)

met_filtered %>% filter(temp > -15) %>% select(temp, elev) %>% summary()

met_filtered %>% filter(temp > -15) -> met_filtered

```
After checking expected temperature range at 29.117, -89.550, it seems that -17.20 celsius is highly unlikely in the southeast coast at this time of year, so we excluded it and anything else below -15 celsius. The new minimum temperature reported is -3 celsius.

# 6: Calculate summary statistics

```{r}

met_filtered %>% arrange(desc(elev)) %>% head(3)

# met_filtered %>% filter(elev == 4113.0) -> met_max_elev

met_filtered %>%
  filter(elev == 4113.0) %>%
  summarise(temp_windsp = cor(temp, wind.sp, use = "complete"),
            temp_day = cor(temp, day, use = "complete"),
            temp_hour = cor(temp, hour, use = "complete"),
            windsp_day = cor(wind.sp, day, use = "complete"),
            windsp_hour = cor(wind.sp, hour, use = "complete")
  )

```

#7: Exploratory graphs

```{r}

met_filtered %>%
  select(elev, temp, wind.sp) %>%
  pivot_longer(cols = everything(), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = val)) +
  geom_histogram(fill = "steelblue", color = "white") +
  facet_wrap(~ var, scales = "free", ncol = 1) +
  theme_minimal()

```

```{r}
met_filtered %>%
  filter(elev == 4113.0) %>%
  leaflet() %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircles(lat =~ lat, lng =~ lon, opacity = 1, fillOpacity = 1, radius = 100)
```

```{r}

met_filtered %>%
  filter(elev == 4113.0) %>%
  mutate(date = ymd_h(paste(year, month, day, hour, sep = " "))) %>%
  relocate(date, .after = min) %>%
  arrange(date) %>%
  select(date, temp, wind.sp) %>%
  pivot_longer(cols = c(temp, wind.sp), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = date, y = val)) +
  geom_line() +
  facet_wrap(~var, scales = "free", ncol = 1) +
  theme_minimal()

```

#8: Ask questions

```{r}

```

