---
title: "Lab_1_OLEA"
author: "Julian Olea"
format: 
  html:
    embed-resources: true
fig-width: 8
fig-height: 6
warning: FALSE
---

```{r Load the packages, output = FALSE}
# install.packages("datasauRus")
library(datasauRus)
library(dplyr)
library(ggplot2)
library(ggpubr)
```


# Question 1:

The help file shows that here are 1846 rows and 3 columns.
We can also see the same information by using dim() as shown below.

```{r}

dim(datasaurus_dozen)

```

The included variables are shown below and identified by using the unique() function

```{r}

unique(datasaurus_dozen$dataset)

```

# Question 2

```{r}

datasaurus_dozen %>%
  filter(dataset %in% "dino") %>%
  {cor(.$x, .$y)}

datasaurus_dozen %>%
  filter(dataset %in% "dino") %>%
  
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  stat_cor(
    method = "pearson",
    label.x.npc = "left",
    label.y.npc = "top",
    parse = FALSE,
    label.sep = "\n",
    size = 4
  )

```

# Question 3

```{r}

datasaurus_dozen %>%
  filter(dataset %in% "star") %>%
  {cor(.$x, .$y)}

datasaurus_dozen %>%
  filter(dataset %in% "star") %>%

ggplot(aes(x = x, y = y)) +
  geom_point() +
   stat_cor(
      method = "pearson",
      label.x.npc = "left",
      label.y.npc = "top",
      parse = FALSE,
      label.sep = "\n",
      size = 4
    )


```

The r value of star is almost identical to that of dino. R of star = -0.063 R of dino = -0.064


# Question 4

```{r}

datasaurus_dozen %>%
  filter(dataset %in% "circle") %>%
  {cor(.$x, .$y)}

datasaurus_dozen %>%
  filter(dataset %in% "circle") %>%

ggplot(aes(x = x, y = y)) +
  geom_point() +
   stat_cor(
      method = "pearson",
      label.x.npc = "left",
      label.y.npc = "top",
      parse = FALSE,
      label.sep = "\n",
      size = 4
    )


```

The r value of circle is slightly less than dino. R of circle = -0.068 R of dino = -0.064

# Question 5

```{r, fig.width = 16, fig.height = 12}

ggplot(datasaurus_dozen, aes(x = x, y = y)) +
  geom_point(aes(color = dataset)) +
   stat_cor(
      method = "pearson",
      label.x.npc = "left",
      label.y.npc = "top",
      parse = FALSE,
      label.sep = "\n",
      size = 4
    ) +
  theme(legend.position = "none") +
  labs(x = "", y = "") +
  facet_wrap(~dataset)

```

# Question 6

```{r}

datasaurus_dozen %>%
  group_by(dataset) %>%
  summarize(r = cor(x, y))

```

